Supplemental Material: Multidimensional analysis and detection of informative features in human brain white matter

### 2021-04-20 reports/manuscript/paper.tex
Adds a section on encoding and decoding.

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Multidimensional analysis and detection of informative features in human brain white matter} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Adam Richie-Halford\textsuperscript{1*},
Jason Yeatman\textsuperscript{2},
Noah Simon\textsuperscript{3},
Ariel Rokem\textsuperscript{4},
\\
\bigskip
\textbf{1} eScience Institute, University of Washington, Seattle, WA, USA
\\
\textbf{2} Graduate School of Education and Division of Developmental and Behavioral Pediatrics, Stanford University, Stanford, CA, USA
\\
\textbf{3} Department of Biostatistics, University of Washington, Seattle, WA, USA
\\
\textbf{4} Department of Psychology, University of Washington, Seattle, WA, USA
\\
\bigskip

@@

samples along the length of each bundle \cite{yeatman2012tract,
colby2012, ODonnell2009-uu},

@@

The present work aims to balance predictive accuracy with descriptive power
\cite{Murdoch2019-ax, Breiman2001-uz} by capitalizing on all of the available
data across all bundles, while also retaining and elucidating spatial
information about the locations that are most informative for discriminative
performance. To do so, we use a linear modeling approach, which aims to
redict phenotypical variance in a group of subjects (or classify group
membership), based on a linear combination of the features estimated with
tract profiling.

@@

Specifically, we demonstrated that the algorithm correctly detects the fact
that ALS, which is a disease of lower motor neurons, is localized to the
cortico-spinal tract. This recapitulates the results of previous analysis of
these same data, using a bundle-of-interest approach
\cite{sarica2017corticospinal}.
As with the original study, it is unclear whether our strong predictive performance based on image properties that would be visible in patient scans or whether it is identifying a new subclinical disease biomarker.

In contrast, for the analysis of biological
age, the coefficients identified by the algorithm are very widely distributed
across many parts of the white matter, mirroring previous results that show a
large and continuous distribution of life-span changes in white matter
properties \cite{yeatman2014lifespan}.
We present age quintile bundle profiles and SGL \protect$\beta$ coefficients for all four datasets in the supplementary material. For age regression, substantial differences between the datasets preclude generalization across all three:
\begin{enumerate*}[%
    label=(\roman*),%
    before=\unskip{: },%
    itemjoin={{, }},%
    itemjoin*={{, and }}]
    \item different acquisition parameters, which can be challenging to harmonize \mbox{\cite{pinto2020harmonization}}
    \item different diffusion models, with DTI for the WH dataset and DKI for the HBN and Cam-CAN datasets
    \item different age ranges and distributions (which is evident in the figure legends for \cref{fig:age-results}), with HBN being a developmental dataset, while WH and Cam-CAN are lifespan maturation datasets
    \item different anatomical extents, with the WH streamlines truncated to remain with the bundle's bounding regions of interest (the default behavior in the legacy \texttt{mAFQ}) and the HBN and Cam-CAN streamlines allowed to retain their full extent from whole-brain tractography (the default behavior in \texttt{pyAFQ}).
\end{enumerate*}
In a biological characteristic that has widespread effects in the brain, validity is difficult to assess \mbox{\cite{lermausabiaga2019replication}, so it is both unsurprising and vexing that we can do so well at predicting within each dataset, and yet do poorly in interpreting informative features across instruments.}

@@

The method is packaged as open-source software called AFQ-Insight that is
openly available at \url{https://github.com/richford/AFQ-Insight}, and
provides a clear API to allow for extensions of the method. The sofware
integrates within a broader automated fiber quantification software
ecosystem: AFQ \cite{yeatman2012tract} and pyAFQ \cite{pyAFQ, kruper2021evaluating}, which extract
tract profile data from raw and processed dMRI datasets, as well as
AFQ-Browser, which visualizes tract profiles data and facilitates sharing of the
results of dMRI studies \cite{yeatman2018browser}. To facilitate
reproducibility and ease use of the software, the results presented in this
paper are also provided in
\url{https://github.com/richford/afq-insight-paper} as a series of Jupyter
notebooks \cite{kluyver2016jupyter}.

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Multidimensional analysis and detection of informative features in human brain white matter} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Adam Richie-Halford\textsuperscript{1*},
Jason Yeatman\textsuperscript{2},
Noah Simon\textsuperscript{3},
Ariel Rokem\textsuperscript{1,4},
\\
\bigskip
\textbf{1} eScience Institute, University of Washington, Seattle, WA, USA
\\
\textbf{2} Graduate School of Education and Division of Developmental and Behavioral Pediatrics, Stanford University, Stanford, CA, USA
\\
\textbf{3} Department of Biostatistics, University of Washington, Seattle, WA, USA
\\
\textbf{4} Department of Psychology, University of Washington, Seattle, WA, USA
\\
\bigskip

@@

samples along the length of each bundle \replaced{\mbox{{\cite{Jones2005-yq, yeatman2012tract, colby2012, ODonnell2009-uu}}}}{\mbox{\cite{yeatman2012tract,
colby2012, ODonnell2009-uu}}},

@@

The present work aims to balance predictive accuracy with descriptive power
\cite{Murdoch2019-ax, Breiman2001-uz} by capitalizing on all of the available
data across all bundles, while also retaining and elucidating spatial
information about the locations that are most informative for discriminative
performance.
 We approach differences between individuals through a ``decoding'' lens:
We predict the phenotypical variance in a group of subjects (or classify group membership), based on a linear combination of the features estimated with tract profiling
This is unlike many previous models of white matter differences (e.g., TBSS \cite{} and connectometry \cite{}) that are ``encoding'' models, where differences between individuals and groups are tested as hypotheses about differences in their brain features (i.e., in their tract profiles; the distinction between ``encoding'' and ``decoding'' and the terms stem from functional MRI literature \cite{Naselaris2011-ss}).

@@

Specifically, we demonstrated that the algorithm correctly detects the fact
that ALS, which is a disease of lower motor neurons, is localized to the
cortico-spinal tract. This recapitulates the results of previous analysis of
these same data, using a bundle-of-interest approach
\cite{sarica2017corticospinal}.
As with the original study, it is unclear whether our strong predictive performance based on image properties that would be visible in patient scans or whether it is identifying a new subclinical disease biomarker.
In contrast, for the analysis of biological
age, the coefficients identified by the algorithm are very widely distributed
across many parts of the white matter, mirroring previous results that show a
large and continuous distribution of life-span changes in white matter
properties \cite{yeatman2014lifespan}.
We present age quintile bundle profiles and SGL \protect$\beta$ coefficients for all four datasets in the supplementary material. For age regression, substantial differences between the datasets preclude generalization across all three:
\begin{enumerate*}[%
    label=(\roman*),%
    before=\unskip{: },%
    itemjoin={{, }},%
    itemjoin*={{, and }}]
    \item different acquisition parameters, which can be challenging to harmonize \mbox{\cite{pinto2020harmonization}}
    \item different diffusion models, with DTI for the WH dataset and DKI for the HBN and Cam-CAN datasets
    \item different age ranges and distributions (which is evident in the figure legends for \cref{fig:age-results}), with HBN being a developmental dataset, while WH and Cam-CAN are lifespan maturation datasets
    \item different anatomical extents, with the WH streamlines truncated to remain with the bundle's bounding regions of interest (the default behavior in the legacy \texttt{mAFQ}) and the HBN and Cam-CAN streamlines allowed to retain their full extent from whole-brain tractography (the default behavior in \texttt{pyAFQ}).
\end{enumerate*}
In a biological characteristic that has widespread effects in the brain, validity is difficult to assess \mbox{\cite{lermausabiaga2019replication}, so it is both unsurprising and vexing that we can do so well at predicting within each dataset, and yet do poorly in interpreting informative features across instruments.}

@@

The method is packaged as open-source software called AFQ-Insight that is
openly available at \url{https://github.com/richford/AFQ-Insight}, and
provides a clear API to allow for extensions of the method. The sofware
integrates within a broader automated fiber quantification software
ecosystem: AFQ \cite{yeatman2012tract} and pyAFQ \cite{kruper2021evaluating}, which extract
tract profile data from raw and processed dMRI datasets, as well as
AFQ-Browser, which visualizes tract profiles data and facilitates sharing of the
results of dMRI studies \cite{yeatman2018browser}. To facilitate
reproducibility and ease use of the software, the results presented in this
paper are also provided in
\url{https://github.com/richford/afq-insight-paper} as a series of Jupyter
notebooks \cite{kluyver2016jupyter}.

### 2021-04-20 reports/manuscript/paper_changes.tex
Adds a section on encoding and decoding.

The present work aims to balance predictive accuracy with descriptive power
\cite{Murdoch2019-ax, Breiman2001-uz} by capitalizing on all of the available
data across all bundles, while also retaining and elucidating spatial
information about the locations that are most informative for discriminative
performance. To do so, we use a linear modeling approach, which aims to
redict phenotypical variance in a group of subjects (or classify group
membership), based on a linear combination of the features estimated with
\replaced{tract profiling}{tractometry}.

>>>>

The present work aims to balance predictive accuracy with descriptive power
\cite{Murdoch2019-ax, Breiman2001-uz} by capitalizing on all of the available
data across all bundles, while also retaining and elucidating spatial
information about the locations that are most informative for discriminative
performance.
\added{ We approach differences between individuals through a ``decoding'' lens:}
We predict the phenotypical variance in a group of subjects (or classify group membership), based on a linear combination of the features estimated with \replaced{tract profiling}{tractometry}
\added{This is unlike many previous models of white matter differences (e.g., TBSS \cite{} and connectometry \cite{}) that are ``encoding'' models, where differences between individuals and groups are tested as hypotheses about differences in their brain features (i.e., in their tract profiles; the distinction between ``encoding'' and ``decoding'' and the terms stem from functional MRI literature \cite{Naselaris2011-ss})}.
